Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-09 paper/paper.tex
Currently stuck in the proof since partitioning isn't helping do case analysis on ifs and applications.

The primary strength of the approach is that abstract interpreters can
be easily derived through a small number of steps from existing
machine models.% such as the CESK machine or the Lazy Krivine machine.
Moreover, since the relationships between abstract machines and
higher-level semantic models---such as definitional
interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational
semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction
semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well
understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
these semantic models to sound program analyzers.

@@

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

@@

\begin{alltt}
;; \(\forall\) (X) ((X \(\to\) Set[X]) \(\to\) (Set[X] \(\to\) Set[X]))
(define ((appl f) s)
  (for/fold ([i (set)])
    ([x (in-set s)])
    (set-union i (f x))))

;; Calculate fixpoint of (appl f).
;; \(\forall\) (X) ((X \(\to\) Set[X]) Set[X] \(\to\) Set[X])
(define (fix f s)
  (let loop ((accum (set)) (front s))
    (if (set-empty? front)
        accum
        (let ((new-front ((appl f) front)))
          (loop (set-union accum front)
                (set-subtract new-front accum))))))
\end{alltt}

@@

;; (State \(\to\) Set[State]) \(\to\) \(\widehat{State}\) \(\to\) \(\widehat{State}\)
(define ((wide-step step) state)
  (match state
    [(cons cs \(\msto\))
     (define ss ((appl step)
                 (for/set ([c cs]) (c->s c \(\msto\)))))
     (set (cons (for/set ([s ss]) (s->c s))
                (join-stores ss)))]))
\end{alltt}

@@

\begin{align*}
\ev{\mexp, \menv, \mkont, \mcntr} \in cs &\Rightarrow \lapprx{\ev{\mexp, \menv,\mkont, \mcntr}}{\msto}{cs} \\
\co{\mkont, \mval} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\co{\mkont, \mval}}{\msto}{cs} \\
\ap{\mval_0, \mval_1, \mkont} \in cs \mbox{ and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \lapprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{cs} \\
\ev{\mexp, \menv, \mkont} \in lcs &\Rightarrow \apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs} \\
\co{\mkont, \mval} \in lcs \mbox { and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\co{\mkont, \mval'}}{\msto}{lcs} \\
\ap{\mval_0, \mval_1, \mkont} \in lcs \mbox { and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \apprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{lcs} \\
\forall lc \in lcs, \lapprx{lc}{\msto}{cs}, \mbox{ and }\forall c \in cs, \apprx{c}{\msto}{lcs}. &\Rightarrow \capprx{(lcs, \msto)}{(cs, \msto)}
\end{align*}

@@

\begin{lemma}[Partition step]
$\forall lc, lc', cs, \msto, \msto'.$ if $\lapprx{lc}{\msto}{cs}$ and $wn(lc,\msto) \machstep wn(lc', \msto')$ then
$\exists cs'. \forall c \in cs. \exists c' \in cs', \msto^c. wn(c,\msto) \machstep wn(c', \msto^c)$ and $\msto' = \bigcup\limits_{c \in cs}{\msto^c}$ and $\lapprx{lc'}{\msto'}{cs'}$.
\end{lemma}
\begin{proof}
By cases on $wn(lc,\msto) \machstep wn(lc', \msto')$:
\begin{itemize}
\item{Case $\ev{\svar{\mvar}, \menv, \mkont} \machstep
            \co{\mkont, \maddr{\menv(\mvar)}, \msto}$: \\
      If $\mvar \in dom(\menv)$ and $\menv(x) \in dom(\msto)$, then $wn(lc, \msto)$ steps
      \begin{itemize} 
       \item{lazily to $\co{\saddr{\menv(\mvar)}, \msto, \mkont}$}
       \item{strictly to each of $\{\co{\mval, \msto, \mkont} \mid \mval \in \msto(\menv(\mvar))\}$}
      \end{itemize}
      By definition of ...}
\item{Case $\ev{\slit{\mlit}, \menv, \msto, \mkont} \machstep
            \co{\mkont, \mlit, \msto}$: \\
      Immeditate.}
\item{Case $\ev{\slam{\mvar}{\mexp}, \menv, \mkont} \machstep
            \co{\mkont, \clos{\mvar, \mexp, \menv}, \msto}$: \\
      Immediate}
\item{Case $\ev[^\mcntr]{\sapp[^\mlab]{\mexp_0}{\mexp_1}, \menv, \mkont} \machstep
            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kar[^\mcntr_\mlab]{\mexp_1, \menv, \maddr}}$: \\
      Where $a, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$
      ...}
\item{Case $\ev[^\mcntr]{\sif[^\mlab]{\mexp_0}{\mexp_1}{\mexp_2}, \menv, \mkont} \machstep
            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kif[^\mcntr]{\mexp_1, \mexp_2, \menv, \maddr}}$: \\
      Where $a, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$            
      ...}
\item{Case $\co{\kmt, \mval, \msto} \machstep \ans{\msto, \mval'}$: \\
      Where $\mval' \in force(\msto, \mval)$
      ...}
\item{Case $\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval}$: \\
      ...}
\item{Case $\co{\kfn[^\mcntr_\mlab]{\mvalx{u}, \maddr}, \mval}$: \\
      ...}
\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \strue}$: \\
      ...}
\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \sfalse}$: \\
      ...}
\item{Case $\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval, \mkont}$: \\
      ...}
\item{Case $\ap[^\mcntr_\mlab]{\mop, \mval, \mkont}$: \\
      ...}
\end{itemize}
\end{proof}

>>>>

The primary strength of the approach is that abstract interpreters can
be easily derived through a small number of steps from existing
machine models. % such as the CESK machine or the Lazy Krivine machine.
Moreover, since the relationships between abstract machines and
higher-level semantic models---such as definitional
interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational
semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction
semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well
understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
these semantic models to sound program analyzers.

@@

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

@@

\begin{alltt}
;; \(\forall\) (X) ((X \(\to\) Set[X]) \(\to\) (Set[X] \(\to\) Set[X]))
(define ((appl f) s)
  (for/fold ([i (set)])
    ([x (in-set s)])
    (\(\cup\) i (f x))))

;; Calculate fixpoint of (appl f).
;; \(\forall\) (X) ((X \(\to\) Set[X]) Set[X] \(\to\) Set[X])
(define (fix f s)
  (let loop ((accum (set)) (front s))
    (if (set-empty? front)
        accum
        (let ((new-front ((appl f) front)))
          (loop (\(\cup\) accum front)
                (new-front . \(\setminus\) . accum))))))
\end{alltt}

@@

;; (State \(\to\) Set[State]) \(\to\) \(\widehat{State}\) \(\to\) \(\widehat{State}\)
(define ((wide-step step) state)
  (match state
    [(cons cs \(\msto\))
     (define ss ((appl step)
                 (for/set ([c cs]) (c->s c \(\msto\)))))
     (set (cons (for/set ([s ss]) (s->c s))
                (\(\sqcup\)-stores ss)))]))
\end{alltt}

@@

\begin{align*}
\ev{\mexp, \menv, \mkont, \mcntr} \in cs &\Rightarrow \lapprx{\ev{\mexp, \menv,\mkont, \mcntr}}{\msto}{cs} \\
\co{\mkont, \mval} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\co{\mkont, \mval}}{\msto}{cs} \\
\ap{\mval_0, \mval_1, \mkont} \in cs \mbox{ and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \lapprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{cs} \\
\ans{\mval'} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\ans{\mval}}{\msto}{cs} \\
\ev{\mexp, \menv, \mkont} \in lcs &\Rightarrow \apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs} \\
\co{\mkont, \mval} \in lcs \mbox { and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\co{\mkont, \mval'}}{\msto}{lcs} \\
\ap{\mval_0, \mval_1, \mkont} \in lcs \mbox { and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \apprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{lcs} \\
\ans{\mval} \in lcs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\ans{\mval'}}{\msto}{lcs} \\
\forall lc \in lcs, \lapprx{lc}{\msto}{cs}, \mbox{ and }\forall c \in cs, \apprx{c}{\msto}{lcs}. &\Rightarrow \capprx{(lcs, \msto)}{(cs, \msto)}
\end{align*}

@@

\begin{lemma}[Partition step]
$\forall lc, lcs', cs, \msto, \msto'.$
 if  $\capprx{(\{lc\},\msto)}{(cs,\msto)}$
 and $\forall lc' \in lcs'. wn(lc,\msto) \machstep wn(lc', \msto')$ then
 $\exists cs'. \forall c \in cs. \exists c' \in cs', \msto^c$
 such that $wn(c,\msto) \machstep wn(c', \msto^c)$
 and       $\msto' = \bigcup\limits_{c \in cs}{\msto^c}$
 and       $\capprx{(lcs',\msto')}{(cs',\msto')}$.
\end{lemma}
\begin{proof}
Let $lc' \in lcs'$ be arbitrary.
By cases on $wn(lc,\msto) \machstep wn(lc', \msto')$:
\begin{itemize}
\item{Case $\ev{\svar{\mvar}, \menv, \mkont, \msto} \machstep \co{\mkont,
    \saddr{\menv(\mvar)}, \msto}$: \\ By definition of $\machstep$,
  $wn(lc,\msto)$ steps strictly to each of $cs' = \{\co{\mval, \mkont} \mid \mval \in \msto(\menv(\mvar))\}$.  By definition of
  $\capprx{}{}$, $cs = \{lc\}$. The stores are the same, and by
  definition of $\lapprx{}{\msto}{}$, $\lapprx{lc'}{\msto}{cs'}$.}
\item{Case $\ev{\slit{\mlit}, \menv, \msto, \mkont, \msto} \machstep
            \co{\mkont, \mlit, \msto}$: \\
      Immeditate.}
\item{Case $\ev{\slam{\mvar}{\mexp}, \menv, \mkont, \msto} \machstep
            \co{\mkont, \clos{\mvar, \mexp, \menv}, \msto}$: \\
      Immediate}
\item{Case $\ev[^\mcntr]{\sapp[^\mlab]{\mexp_0}{\mexp_1}, \menv, \mkont, \msto} \machstep
            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kar[^\mcntr_\mlab]{\mexp_1, \menv, \maddr}, \msto'}$: \\
      Where $\maddr, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$ \\
      Immediate.}
\item{Case $\ev[^\mcntr]{\sif[^\mlab]{\mexp_0}{\mexp_1}{\mexp_2}, \menv, \mkont, \msto} \machstep
            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kif[^\mcntr]{\mexp_1, \mexp_2, \menv, \maddr}, \msto'}$: \\
      Where $\maddr, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$ \\
      Immediate.}
\item{Case $\co{\kmt, \mval, \msto} \machstep \ans{\msto, \mval'}$: \\
      Where $\mval' \in force(\msto, \mval)$ \\
      By definition of $\capprx{}{}$, $cs = \{\co{\kmt, \mval'} \mid \mval \in force(\msto, \mval)\}$.
      By definition of $\machstep$, $wn(lc,\msto)$ steps strictly to
      each of $cs' = \{\ans{\mval}\}$. The stores are the same.
      By definition of $\lapprx{}{\msto}{}$, $\forall lc' \in lcs'. \lapprx{lc'}{\msto}{cs'}$.
      By definition of $\apprx{}{\msto}{}$, $\forall c' \in cs'. \apprx{c'}{\msto}{lcs'}$.
      Thus $\capprx{(lcs',\msto)}{(cs',\msto)}$.}
\item{Case $\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval, \msto} \machstep
            \ev[^\mcntr]{\mexp, \menv, \kfn[^\mcntr_\mlab]{\mval',\maddr}, \msto}$: \\
      Where $v' \in force(\msto, \mval)$. \\
      By definition of $\apprx{}{\msto}{}$, 
        $cs = \{\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval'} \mid \mval \in force(\msto, \mval)\}$.
      By definition of $\machstep, \lapprx{}{\msto}{}$,
        $cs' = \{\ev[^\mcntr]{\mexp, \menv, \kfn[^\mcntr_\mlab]{\mval',\maddr}}
            \mid \co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval'} \in cs\}$.
      The stores are the same, and the previous imply $\capprx{(lcs',\msto)}{(cs',\msto)}$.}
\item{Case $\co{\kfn[^\mcntr_\mlab]{\mvalx{u}, \maddr}, \mval, \msto} \machstep
            \ap{\mvalx{u}, \mval, \mkont, \msto}$: \\
      Where $\mkont \in \msto(\maddr)$. \\
      Immediate.}
\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
            \ev{\mexp_i, \menv, \mkont, \msto}$: \\
      Where $\mkont \in \msto(\maddr)$. \\
      By cases on $\{\strue, \sfalse\} \cap force(\msto, \mval)$: \\
 \begin{itemize}
  \item{Case $\{\strue, \sfalse\}$:}
  \item{Case $\{\strue\}$:}
  \item{Case $\{\sfalse\}$:}
 \end{itemize}
 ...}
%% \item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
%%             \ev{\mexp_1, \menv, \mkont, \msto}$: \\
%%       Where $\sfalse \in force(\msto, \mval)$ and $\mkont \in \msto(\maddr)$. \\
%%       ...}
\item{Case $\ap[^\mcntr_\mlab]{\mvalx{u}, \mval, \mkont,\msto} \machstep
            \ev[^{\mcntr'}]{\mexp, \menv', \mkont, \msto'}$: \\
      {\bf BREAKDOWN}
      Where $\clos{\mvar, \mexp, \menv} \in force(\msto, \mvalx{u})$
        and $\menv', \msto', \mcntr' = \mathit{bind}^\mcntr_\mlab(\menv, \msto, \mvar, \mval)$
      ...}
\item{Case $\ap[^\mcntr_\mlab]{\mvalx{u}, \mval, \mkont,\msto} \machstep
            \co{\mval'', \mkont, \msto}$: \\
      {\bf BREAKDOWN}
      Where $\mop \in force(\msto, \mvalx{u})$, $\mval' \in force(\msto, \mval)$
        and $\mval'' \in\interpdelta(\mop,\mval')$. \\
      ...}
\end{itemize}
\end{proof}
